When writing a dissertation about historical contingency, it is impossible to avoid thinking about the events in my own life, both big and small, that have brought me to this point. 
Among all of the decisions, failures, and chance events that potentiated this path, the most important factors are the plethora of amazing people whose paths have crossed my own.
Here I wish to thank them.
Fully encapsulating my gratitude within this text is impossible, but I will try my best. 

To those who seemingly never tire: Barbara Bloemers, Vinny Mattison, Dr. Elise Zipkin, Dr. Sandeep Kulkarni, and Dr. Eric Torng, you each deserve sainthood for your demonstrations of patience and tenacity in pushing for us graduate students. 

To those most likely to read this dissertation: Dr. Chris Adami, Dr. Wolfgang Banzhaf, and Dr. Rob Pennock, your patience and guidance have helped shape me into a scientist. 

To those who led the way: Dr. Emily Dolson, Dr. Josh Nahum, Dr. Alex Lalejini, Alexa Lalejini, Dr. Matthew Moreno, Dr. Jose Hernandez, Dr. Anselmo Pontes, Dr. Anya Vostinar, Dr. Mike Wiser, Dr. Vinny Ragusa, Dr. Jory Schossau, Santiago Rodriguez Papa, Keron Greene, and Dr. Nkrumah Grant, you  welcomed me with open arms and offered friendship, guidance, and much needed reassurance every step of the way. 

To those who kept me sane: Dr. Acacia Ackles and Kate Skocelas, without the two of you I would not have made it through the isolation of a global pandemic. 

To those who carried me through the finish: Sydney Leither, Max Foreback, Shakiba Shahbandegan, and Karen Suzue, the countless nights of laughter have made these last couple years some of my best. 
Your unwavering support and wisdom beyond your years keep me in grateful awe. 

To the one who is uncategorizable: Cliff Bohm, who always has an idea to debate, a paper to collaborate on, and a warm dinner for good company. 

To that good company: Chris Reeves, Kira Chan, Lucky Voutour, Ariel Cooper, and the rest of the crew, your comradery was always a highlight of my week. 

To those who made Monday mornings tolerable: the EEB coworking crew, there is no one I would rather work in silence around. You are truly the most supportive group of people I have ever met. 

To those who let me learn how to teach: my undergraduate mentees Daniel Junghans, Nick Lloyd, Caroline Gormerly, Dylan Rainbow, Lanea Rohan, Aria Killebrew-Bruehl, Valdine Pegy Tchinda, and dozen of students across classes and workshops, thank you for your patience and in teaching me more than I taught you. 

To those who are always there: Dr. Ronnie Emmons, Logan Boggs, Dariel Marine, Dr. Gabe Howard, and Jason Montavon, your companionship and endless laughs are truly appreciated. 

To the one who pushed me: Andrew Polanco, you single-handedly convinced me to aim for what I really wanted and refused to let me falter. 

To those who demonstrated the role: my instructors and colleagues at Shawnee, it was your passion for educating and your kind hearts that led me to believe I could maybe pull this off. 

To the friends who have truly become family: Alex Bellomy, who continuously raises the bar for what it means to be someone's best friend; and Ashlyn Bellomy, who has demonstrated that true friends do not require a lifetime of history.

To the one who took a chance on me: Dr. Charles Ofria, my advisor who foolishly accepted a clueless new graduate student six years ago.
Through a wild journey, you have shaped me into a scientist and educator, never showing the slightest hesitation to help me or just anyone around you.
While I am sad to see this journey end, I am so happy to call you a friend at the end of it.

To those that gave me unwavering support: Martha, Jim, and Marty Ferguson and Jeff Adkins, you have never forgotten to say ``I'm proud of you'' when I needed it most. 

To the one taken too soon: Debbie Adkins, you taught me the most important lessons: not to take life too seriously and to always love with all your heart. 

To the one who changed my life: Ethan Ferguson, your big heart and determination continuously fill me with pride.
You may be the younger brother, but I am the one learning from you. 

To the ones who made it possible: Donnie and Alisha Ferguson, you have always pushed to give me the opportunities you did not have. 
You taught me the true value in ``we are here to support you''. 
You also taught me how to work hard, laugh harder, and make the most of life. 
Everything I have done and will do in this life are thanks to the sacrifices you have made.

To those mentioned and those not: from the bottom of my heart, thank you.